A Mechanised Library for Generic (Co)Programming

### 2024-02-27 paper/paper.tex
Hylo stuff

Algebras on container extensions are defined as expected. The carrier of the
algebra is \coq{A}, and the operations of the algebra are morphisms
of the following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
Given a container extension \coq{F}, we define \emph{two}
(isomorphic) initial algebras. We will show the need for the second initial
algebra in the following section. Our fist initial algebra is defined as the
straightforward least fixpoint of the container extension \coq{F}:
\begin{coqcode}
Inductive LFix  : Type := LFix_in { LFix_out : App F LFix }.
\end{coqcode}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words,
we would need to provide an equivalence for \coq{LFix}, but the
only way to do this is to define such equivalence simultaneously to the
definition for  \coq{LFix}.  We could have overcame this
circularity by defining  \coq{LFix} independently of container
extensions, and define later the necessary
\coq{in}/\coq{out} functions to convert between
\coq{LFix} and \coq{App}. However, this extra layer
interferes with code extraction, in that it requires destructing/constructing
subsequent \coq{LFix} and  \coq{App}, which Coq is not
always able to inline.

We provide a custom induction scheme for initial algebras:
\begin{coqcode}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{coqcode}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \coq{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \coq{LFixR}.
\begin{coqcode}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{coqcode}
From such initial algebras, we define \emph{catamorphisms} by recursively
traversing their structure. As usual, a catamorphism takes an
\coq{F}-algebra with carrier \coq{A}, and produces a
function from \coq{LFix F} to \coq{A}.
\begin{coqcode}
Definition cata_f `{setoid A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkApp sx (fun e => f (kx e)))
  end.
\end{coqcode}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{coqcode}
Definition cata `{eA : setoid A} : Alg F A ~> LFix ~> A := ...
\end{coqcode}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \coq{f =e g \o fmap f \o l_out}).
The proof is by induction on the input to \coq{f} and
\coq{cata g}.
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{coqcode}

@@

Terminal coalgebras are defined analogously to initial algebras:
\begin{coqcode}
CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
\end{coqcode}
Their equivalence relation, and anamorphisms, and universal properties are
easy to encode:
\begin{coqcode}
  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...

  Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
This definition, however, prevents us from composing catamorphisms and
anamorphisms. To address this issue, we consider another alternative definition
for initial $F$-algebras. In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Catamorphisms from this initial algebra are denoted by \coq{ccata},
and they can be used to define the isomorphism between 
\coq{{x : GFix F | FinF x}} and
\coq{LFix F}. In particular: 
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
the supplementary material and Coq code to see how we encode this. In this
paper, however, we focus on the fact that, anamorphisms produce finite trees,
then their return type can also be the \emph{least} fixed point of the
container extension. We encode in Coq the notion of \emph{recursive
anamorphisms} and \emph{recursive coalgebras}~\cite{TODO}.  Recursive
coalgebras are coalgebras whose anamorphism produces finite trees.  The
following predicate captures this fact. The following predicate states that
\coq{h} can only be applied finitely many times:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
We show how this predicate is equivalent to the anamorphism of such coalgebras
producing finite trees:
\begin{coqcode}
Lemma ana_rec_term `{setoid A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{coqcode}
Recursive anamorphisms are implemented in Coq recursively on the structure of
the proof that they can only be applied finitely many times
(\coq{RecF}).  Note that such proofs will be erased when extracting
code to OCaml. This is standard practice when building general recursive
functions in Coq\footnote{\dcas{This technique had a name in the community, but
I cannot remember. Something predicates}}. The core definitions are the
following:
\begin{coqcode}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).

@@

Given an $F$-algebra and $F$-coalgebra, a solution to the equation $f = a \circ
F\; f \circ c$ is known as a hylomorphism. Unfortunately, this definition
cannot be accepted by Coq, since there is no structurally smaller argument to
do recursion, and there is no corecursive definition that is guaranteed to be
guarded, due to the use of an arbitrary algebra $a$. Furthermore, solutions to
the hylomorphism equation are \emph{not} guaranteed to be unique, or to even
exist. To address these issues, we mechanise \textbf{recursive hylomorphisms}
which are guaranteed to have a unique solution to the hylomorphism equation.

If $c$ is a recursive coalgebra, then the equation $f = a \circ F\; f \circ c$ 
does have a unique solution.
We show this by first providing the definition
for hylomorphisms in Coq, and then proving that any function satisfying this
equation is a hylomorphism. We call these hylomorphisms with a recursive
coalgebra, \textbf{recursive hylomorphisms}.
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A) 
  : forall (x : A), RecF c x -> B := fix f x H 
  := match c x as h0 return
       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with
     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
The definition of recursive hylomorphisms is recursive on the proof that the
anamorphism can only be applied finitely many times on any input. We again use
\coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls.  As we did with catamorphisms and recursive anamorphisms,
we prove that \coq{hylo_def} is respectful, and use this proof to
built the corresponding higher-order proper morphism:
\begin{coqcode}
Definition hylo : Alg F B ~> RCoalg F A ~> A ~> B := ...
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to
the hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
This result can be used to prove hylomorphism laws in a way that is analogous
to pen-and-paper program calculation techniques. We show the rule for 
hylomorphism fusion as an example:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) 
  (g2 : Alg F C) (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
  : f2 \o hylo g1 h1 =e hylo g2 h1.
Proof.
  (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
  rewrite hylo_univ.
  (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
  rewrite fmap_comp.
  (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
  rewrite compA.
  (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
  rewrite <- E2.
  (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
  rewrite <- compA.
  (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
  rewrite <- compA.
  (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
  rewrite (compA g1).
  (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
  rewrite <- hylo_unr.
  (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
  reflexivity.
Qed.
\end{coqcode}
Using the hylo fusion law, we can prove the well-known \emph{deforestation
optimisation}, where two consecutive recursive computations, one that builds a
data structure, and another one that consumes it, can be fused together into a
single recursive definition. This, in turn, allows us to prove that a recursive
hylomorphism is the composition of a catamorphism and a recursive anamorphism.
\begin{coqcode}
Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.

>>>>

Algebras on container extensions are defined as expected. The carrier of the
algebra is \coq{A}, and the operations of the algebra are morphisms
of the following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
Given a container extension \coq{F}, we define \emph{two}
(isomorphic) initial algebras. We will show the need for the second initial
algebra in the following section. Our fist initial algebra is defined as the
straightforward least fixpoint of the container extension \coq{F}:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words,
we would need to provide an equivalence for \coq{LFix}, but the
only way to do this is to define such equivalence simultaneously to the
definition for  \coq{LFix}. Such definition would be rejected by Coq.
There are other approaches that can circumvent this problem, by separating
the definition of least fixed points from the proofs that the functions that
get the contained elements are proper morphisms. But this leads to longer,
unnecessarily complicated proofs.  Furthermore, this extra layer interferes
with code extraction, in that it requires destructing/constructing subsequent
\coq{LFix} and  \coq{App}, and Coq may not always able to inline them.

%%\dcas{Boring stuff}
%The definition of \coq{LFix} in terms of \coq{App} requires us to define our
%own custom induction scheme for initial algebras:
%\begin{coqcode}
%Lemma LFix_ind [P : LFix -> Prop]
%  : (forall x : App F LFix,
%        (forall e : Pos (shape x), P (cont x e)) ->
%        P (LFix_in x))
%    -> forall l : LFix, P l.
%\end{coqcode}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \coq{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \coq{LFixR}.
\begin{coqcode}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{coqcode}

Least fixed points of container extensions come with two functions that are
defined as wrappers around constructor \coq{LFix_in} and projection
\coq{LFix_out}.
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F
l_out : LFix F ~> App F (LFix F)
\end{coqcode}

We define \emph{catamorphisms} by recursively traversing their structure of the
input \coq{LFix}. As usual, a catamorphism takes an \coq{F}-algebra with
carrier \coq{A}, and produces a function from \coq{LFix F} to \coq{A}.
\begin{coqcode}
Definition cata_f (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkApp sx (fun e => f (kx e)))
  end.
\end{coqcode}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{coqcode}
Definition cata : Alg F A ~> LFix ~> A := ...
\end{coqcode}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \coq{f =e g \o fmap f \o l_out}).
The proof is by induction on the input to \coq{f} and
\coq{cata g}.
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{coqcode}

@@

Terminal coalgebras are defined analogously to initial algebras (see Figure~\ref{fig:term_coalg}).
\begin{figure}
\begin{coqcode}
Context `(F : Cont Sh Po).
CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
Definition GFixR : GFix -> GFix -> Prop :=...
Instance GFix_setoid : setoid GFix := ...GFixR...
Definition g_in : App F GFix ~> GFix := ...
Definition g_out : GFix ~> App F GFix := ...
\end{coqcode}
\caption{Terminal Coalgebras}\label{fig:term_coalg}
\end{figure}
Their equivalence relation, and anamorphisms, and universal properties are
easy to encode, in an analogous way to catamorphisms and initial algebras:
\begin{coqcode}
  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...

  Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
This definition, however, prevents us from composing catamorphisms and
anamorphisms: we cannot mix \coq{GFix} and \coq{LFix}.
To address this issue, we consider another alternative definition
for initial $F$-algebras. In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Catamorphisms from this initial algebra are denoted by \coq{ccata},
and they can be used to define the isomorphism between 
\coq{{x : GFix F | FinF x}} and
\coq{LFix F}. In particular: 
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
the supplementary material and Coq code to see how we encode this. In this
paper, however, we focus on the fact that, if anamorphisms produce finite
trees, then their return type can also be the \emph{least} fixed point of the
container extension. We encode in Coq the notion of \emph{recursive
anamorphisms} and \emph{recursive coalgebras}~\cite{TODO}.  Recursive
coalgebras are coalgebras whose anamorphism produces finite trees.  The
following predicate captures this fact, \coq{h} is recursive if
\coq{h} can only be applied finitely many times:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
This predicate is equivalent to the proposition that the anamorphism of such
coalgebras produce only finite trees:
\begin{coqcode}
Lemma ana_rec_term `{setoid A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{coqcode}
Recursive anamorphisms are implemented in Coq recursively on the structure of
the proof that they can only be applied finitely many times (\coq{RecF}).  Note
that such proofs will be erased when extracting code to OCaml, and this is
standard practice when building general recursive functions in
Coq\footnote{\dcas{This technique had a name in the community, but I cannot
remember. Something predicates}}. The core definitions are the following:
\begin{coqcode}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).

@@

Given an $F$-algebra and $F$-coalgebra, a solution to the equation $f = a \circ
F\; f \circ c$ is known as a hylomorphism. Unfortunately, this definition
cannot be accepted by Coq, since there is no structurally smaller argument to
do recursion, and there is no corecursive definition that is guaranteed to be
syntactically guarded (as required by Coq), due to the use of an arbitrary
algebra $a$. Furthermore, solutions to
the hylomorphism equation are \emph{not} guaranteed to be unique, or to even
exist. To address these issues, we mechanise \textbf{recursive hylomorphisms}
which are guaranteed to have a unique solution to the hylomorphism equation.

If $c$ is a recursive coalgebra, then the equation $f = a \circ F\; f \circ c$
does have a unique solution.  We show this by first providing the definition
for recursive hylomorphisms in Coq, and then proving that any function
satisfying this equation is a hylomorphism. We call these hylomorphisms with a
recursive coalgebra, \textbf{recursive hylomorphisms}.
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A) 
  : forall (x : A), RecF c x -> B := fix f x H 
  := match c x as h0 return
       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with
     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
The definition of recursive hylomorphisms is recursive on the proof that the
anamorphism can only be applied finitely many times on any input. We again use
\coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls.  As we did with catamorphisms and recursive anamorphisms,
we prove that \coq{hylo_def} is respectful, and use this proof to
built the corresponding higher-order proper morphism:
\begin{coqcode}
Definition hylo : Alg F B ~> RCoalg F A ~> A ~> B := ...
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to
the hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
This result can be used to prove hylomorphism laws in a way that is analogous
to pen-and-paper program calculation techniques. We show the rule for 
hylomorphism fusion as an example:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) 
  (g2 : Alg F C) (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
  : f2 \o hylo g1 h1 =e hylo g2 h1.
Proof.
  rewrite hylo_univ, fmap_comp compA <- E2 <- compA <- compA (compA g1) <- hylo_unr
  reflexivity.
Qed.
  (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
  (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
  (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
  (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
  (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
  (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
  (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
  (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
  (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
\end{coqcode}
Using the hylo fusion law, we can prove the well-known \emph{deforestation
optimisation}, where two consecutive recursive computations, one that builds a
data structure, and another one that consumes it, can be fused together into a
single recursive definition. This, in turn, allows us to prove that a recursive
hylomorphism is the composition of a catamorphism and a recursive anamorphism.
\begin{coqcode}
Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.
